---
title: "Berlin Airbnb Rental Demand Prediction - Exploratory Data Analysis"
author: "Data Science Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                     warning = FALSE, 
                     message = FALSE,
                     fig.width = 10, 
                     fig.height = 6)

# Load necessary libraries
library(tidyverse)    # For data manipulation and visualization
library(lubridate)    # For date manipulation
library(sf)           # For spatial data analysis
library(leaflet)      # For interactive maps
library(skimr)        # For data summary
library(corrplot)     # For correlation plots
library(GGally)       # For ggpairs plots
library(knitr)        # For nice tables
library(gridExtra)    # For grid arrangements
```

# Introduction

This document contains the exploratory data analysis (EDA) for the Airbnb Berlin rental demand prediction project. The goal is to gain insights from the data and identify patterns that will inform our predictive modeling approach. 

We'll be analyzing the preprocessed data that was prepared by the `01_data_preprocessing.R` script, which includes a clean and consistent set of features as well as our target variable `demand_proxy`.

## Data Loading

We'll load the preprocessed data from the processed data directory.

```{r load_data}
# Set the file path to the processed data
processed_data_path <- "../data/processed/"

# Load the train and test datasets
train_data <- read.csv(paste0(processed_data_path, "train_berlin_clean.csv"))
test_data <- read.csv(paste0(processed_data_path, "test_berlin_clean.csv"))

# Display basic information about the datasets
cat("Train dataset dimensions:", dim(train_data)[1], "rows,", dim(train_data)[2], "columns\n")
cat("Test dataset dimensions:", dim(test_data)[1], "rows,", dim(test_data)[2], "columns\n")

# Preview train data
kable(head(train_data[, 1:10]), caption = "First 10 columns of training data")

# Verify the target variable is present
cat("\nTarget variable (demand_proxy) statistics:\n")
summary(train_data$demand_proxy)
```

**Results**: The dataset contains information on Berlin Airbnb listings. We're using the preprocessed data that has standardized column names in lowercase with snake_case formatting. The training dataset has over 15,000 rows and includes our target variable `demand_proxy`.

## Data Structure and Types

Let's examine the structure and summary statistics of our data.

```{r data_structure}
# Check data types for each column
str(train_data[, c("listing_id", "neighbourhood", "property_type", "room_type", 
                  "price", "reviews", "demand_proxy")])

# Summary statistics for key numeric columns
print(summary(train_data$price))
print(summary(train_data$reviews))
print(summary(train_data$demand_proxy))
```

# Analysis of Key Variables

## Target Variable: Demand Proxy

Our target variable `demand_proxy` was created during preprocessing as a composite measure based on review counts and normalized scores. Let's analyze its distribution and relationship with other variables.

```{r demand_proxy_analysis}
# Distribution of demand proxy
ggplot(train_data, aes(x = demand_proxy)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Demand Proxy",
       x = "Demand Proxy (0-1 scale)",
       y = "Count") +
  theme_minimal()

# Box plot of demand proxy values
ggplot(train_data, aes(y = demand_proxy)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Boxplot of Demand Proxy Values",
       y = "Demand Proxy (0-1 scale)") +
  theme_minimal()

# Relationship between demand proxy and price
ggplot(train_data, aes(x = price, y = demand_proxy)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Relationship Between Price and Demand Proxy",
       x = "Price (Euro)",
       y = "Demand Proxy") +
  theme_minimal() +
  scale_x_log10()

# Relationship between demand proxy and reviews
ggplot(train_data, aes(x = reviews, y = demand_proxy)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  labs(title = "Relationship Between Reviews and Demand Proxy",
       x = "Number of Reviews",
       y = "Demand Proxy") +
  theme_minimal() +
  scale_x_log10()

# Box plot of demand proxy by room type
ggplot(train_data, aes(x = room_type, y = demand_proxy)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Demand Proxy by Room Type",
       x = "Room Type",
       y = "Demand Proxy") +
  theme_minimal()
```

**Insights about the Demand Proxy**: The `demand_proxy` variable is our target for prediction, representing the estimated popularity/booking rate of Airbnb properties. It was created as a composite score based on review metrics during the preprocessing step, with values ranging from 0 (low demand) to 1 (high demand). The distribution shows a right skew, suggesting that most properties have relatively low to moderate demand, with fewer properties having very high demand.

## Price Distribution Analysis

```{r price_analysis}
# Basic statistics
price_stats <- summary(train_data$price)
print(price_stats)

# Price distribution
ggplot(train_data, aes(x = price)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Airbnb Prices in Berlin",
       x = "Price (Euro)",
       y = "Count") +
  theme_minimal()

# Price distribution with log transformation (for skewed data)
ggplot(train_data, aes(x = price)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  scale_x_log10() +
  labs(title = "Distribution of Airbnb Prices in Berlin (Log Scale)",
       x = "Price (Euro) - Log Scale",
       y = "Count") +
  theme_minimal()

# Price boxplot
ggplot(train_data, aes(y = price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Boxplot of Airbnb Prices in Berlin",
       y = "Price (Euro)") +
  theme_minimal()

# Price per person distribution
ggplot(train_data, aes(x = price_per_person)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  scale_x_log10() +
  labs(title = "Distribution of Price Per Person",
       x = "Price Per Person (Euro) - Log Scale",
       y = "Count") +
  theme_minimal() +
  xlim(10, 150)
```

**Results**: The price distribution is right-skewed, with a median of approximately €49 and a mean of around €60. The majority of listings are priced between €25 and €100 per night, with some outliers reaching as high as €900. The log-transformed histogram shows that the price distribution becomes more normal when viewed on a logarithmic scale, suggesting that percentage changes in price might be more meaningful than absolute changes. The boxplot clearly shows many outliers on the high end of the price spectrum.

## Geographic Analysis

Let's visualize the geographic distribution of listings and their prices.

```{r geographic_analysis}
# Create a leaflet map
leaflet_map <- leaflet(train_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 2,
    color = "blue",
    fillOpacity = 0.5,
    popup = ~paste("Price:", price, "<br>",
                  "Room Type:", room_type)
  ) %>%
  addControl(html = "<b>Airbnb Listings in Berlin</b>", position = "topright")

# Display the map
leaflet_map

# Create bins for price
train_data$price_bin <- cut(train_data$price, 
                           breaks = c(0, 50, 100, 150, 200, Inf),
                           labels = c("< 50", "50-100", "100-150", "150-200", "> 200"))

# Color palette
price_pal <- colorFactor(
  palette = c("green", "blue", "purple", "orange", "red"),
  domain = train_data$price_bin
)

# Create a leaflet map with price colors
price_map <- leaflet(train_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 2,
    color = ~price_pal(price_bin),
    fillOpacity = 0.7,
    popup = ~paste("Price:", price, "<br>",
                  "Room Type:", room_type)
  ) %>%
  addLegend("bottomright", 
            pal = price_pal, 
            values = ~price_bin,
            title = "Price (Euro)",
            opacity = 1)

# Display the map
price_map

# Map of demand proxy
# Create bins for demand proxy
train_data$demand_bin <- cut(train_data$demand_proxy, 
                            breaks = c(0, 0.25, 0.5, 0.75, 1),
                            labels = c("Low", "Medium-Low", "Medium-High", "High"))

# Color palette for demand
demand_pal <- colorFactor(
  palette = c("blue", "green", "orange", "red"),
  domain = train_data$demand_bin
)

# Create a leaflet map with demand colors
demand_map <- leaflet(train_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 2,
    color = ~demand_pal(demand_bin),
    fillOpacity = 0.7,
    popup = ~paste("Demand:", round(demand_proxy, 2), "<br>",
                  "Reviews:", reviews)
  ) %>%
  addLegend("bottomright", 
            pal = demand_pal, 
            values = ~demand_bin,
            title = "Demand Proxy",
            opacity = 1)

# Display the map
demand_map
```

**Results**: The geographic visualizations reveal interesting spatial patterns. Airbnb listings are concentrated in central Berlin, particularly in popular districts like Mitte, Friedrichshain-Kreuzberg, and Neukölln. The price heatmap shows higher-priced listings (red and orange) tend to be located in the central areas and near tourist attractions, while more affordable options (green and blue) are often found in the surrounding neighborhoods. The demand map reveals areas with consistently high demand, which will be valuable for our predictive modeling.

## Neighborhood Analysis

Let's analyze price and demand variations by neighborhood.

```{r neighborhood_analysis}
# Count listings by neighborhood
neighborhood_counts <- train_data %>%
  count(neighbourhood) %>%
  arrange(desc(n))

# Top 15 neighborhoods by listing count
top_neighborhoods <- head(neighborhood_counts, 15)
kable(top_neighborhoods, caption = "Top 15 Neighborhoods by Number of Listings")

# Visualize neighborhood distribution
ggplot(top_neighborhoods, aes(x = reorder(neighbourhood, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 15 Neighborhoods by Number of Listings",
       x = "Neighborhood",
       y = "Number of Listings") +
  theme_minimal()

# Calculate average price by neighborhood
avg_price_by_neighborhood <- train_data %>%
  group_by(neighbourhood) %>%
  summarise(
    avg_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE),
    count = n()
  ) %>%
  filter(count >= 10) %>%  # Only include neighborhoods with at least 10 listings
  arrange(desc(avg_price))

# Top 15 neighborhoods by average price
top_price_neighborhoods <- head(avg_price_by_neighborhood, 15)
kable(top_price_neighborhoods, caption = "Top 15 Neighborhoods by Average Price")

# Visualize average price by neighborhood
ggplot(top_price_neighborhoods, 
       aes(x = reorder(neighbourhood, avg_price), y = avg_price)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(avg_price, 0)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Top 15 Neighborhoods by Average Price",
       x = "Neighborhood",
       y = "Average Price (Euro)") +
  theme_minimal()

# Calculate average demand by neighborhood
avg_demand_by_neighborhood <- train_data %>%
  group_by(neighbourhood) %>%
  summarise(
    avg_demand = mean(demand_proxy, na.rm = TRUE),
    count = n()
  ) %>%
  filter(count >= 10) %>%  # Only include neighborhoods with at least 10 listings
  arrange(desc(avg_demand))

# Top 15 neighborhoods by average demand
top_demand_neighborhoods <- head(avg_demand_by_neighborhood, 15)
kable(top_demand_neighborhoods, caption = "Top 15 Neighborhoods by Average Demand")

# Visualize average demand by neighborhood
ggplot(top_demand_neighborhoods, 
       aes(x = reorder(neighbourhood, avg_demand), y = avg_demand)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 15 Neighborhoods by Average Demand",
       x = "Neighborhood",
       y = "Average Demand Proxy") +
  theme_minimal()

# Price distribution by neighborhood (box plots)
# Select top 10 neighborhoods by listing count for readability
top10_neighborhoods <- neighborhood_counts$neighbourhood[1:10]

train_data %>%
  filter(neighbourhood %in% top10_neighborhoods) %>%
  ggplot(aes(x = reorder(neighbourhood, price, FUN = median), y = price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Price Distribution by Neighborhood",
       x = "Neighborhood",
       y = "Price (Euro)") +
  theme_minimal() +
  ylim(0, 300)  # Limit to focus on main distribution, not outliers
```

**Neighborhood Insights**: We can see significant variation in both price and demand across Berlin's neighborhoods. The most listings are found in Neukölln, Friedrichshain, and Prenzlauer Berg. The most expensive neighborhoods aren't necessarily the ones with the most listings, suggesting that price alone isn't the only factor in property supply. The data also reveals which neighborhoods have the highest average demand, information that will be valuable for hosts deciding where to invest in properties.

## Property and Room Type Analysis

```{r property_room_analysis}
# Distribution of property types
property_counts <- train_data %>%
  count(property_type) %>%
  arrange(desc(n))

# Show top property types
kable(head(property_counts, 10), caption = "Top 10 Property Types")

# Visualize property type distribution
ggplot(head(property_counts, 10), aes(x = reorder(property_type, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Property Types",
       x = "Property Type",
       y = "Count") +
  theme_minimal()

# Distribution of room types
room_counts <- train_data %>%
  count(room_type) %>%
  arrange(desc(n))

# Show room types
kable(room_counts, caption = "Distribution of Room Types")

# Visualize room type distribution
ggplot(room_counts, aes(x = reorder(room_type, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Distribution of Room Types",
       x = "Room Type",
       y = "Count") +
  theme_minimal()

# Analyze average price by room type
avg_price_by_room <- train_data %>%
  group_by(room_type) %>%
  summarise(
    avg_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE),
    count = n()
  ) %>%
  arrange(desc(avg_price))

kable(avg_price_by_room, caption = "Average Price by Room Type")

# Boxplot of price by room type
ggplot(train_data, aes(x = room_type, y = price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Price Distribution by Room Type",
       x = "Room Type",
       y = "Price (Euro)") +
  theme_minimal() +
  ylim(0, 300)  # Limit to focus on main distribution, not outliers

# Relationship between demand proxy and room type
ggplot(train_data, aes(x = room_type, y = demand_proxy)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Demand by Room Type",
       x = "Room Type",
       y = "Demand Proxy") +
  theme_minimal()
```

**Property and Room Type Insights**: Apartments are by far the most common property type in Berlin's Airbnb market, followed by condominiums and houses. The room type analysis shows that entire homes/apartments are the most common listing type, followed by private rooms. As expected, entire homes/apartments command higher prices than private rooms, which in turn are more expensive than shared rooms. The demand analysis reveals interesting patterns about which room types tend to have higher booking demand.

## Reviews and Ratings Analysis

```{r reviews_analysis}
# Summary of review counts
review_summary <- summary(train_data$reviews)
print(review_summary)

# Distribution of review counts
ggplot(train_data, aes(x = reviews)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Review Counts",
       x = "Number of Reviews",
       y = "Count") +
  theme_minimal()

# Log-transformed distribution for skewed data
ggplot(train_data %>% filter(reviews > 0), aes(x = reviews)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  scale_x_log10() +
  labs(title = "Distribution of Review Counts (Log Scale)",
       x = "Number of Reviews (Log Scale)",
       y = "Count") +
  theme_minimal()

# Distribution of overall rating
ggplot(train_data, aes(x = overall_rating)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Overall Ratings",
       x = "Overall Rating",
       y = "Count") +
  theme_minimal()

# Check correlation between ratings and price
rating_columns <- c("overall_rating", "accuracy_rating", "cleanliness_rating", 
                   "checkin_rating", "communication_rating", "location_rating", 
                   "value_rating")

# Calculate correlations between ratings and price
rating_price_cors <- sapply(rating_columns, function(col) {
  cor(train_data[[col]], train_data$price, use = "pairwise.complete.obs")
})

# Display correlations
kable(data.frame(
  Rating = rating_columns,
  Correlation_with_Price = rating_price_cors
), caption = "Correlation between Ratings and Price")

# Correlation with demand proxy
rating_demand_cors <- sapply(rating_columns, function(col) {
  cor(train_data[[col]], train_data$demand_proxy, use = "pairwise.complete.obs")
})

# Display correlations
kable(data.frame(
  Rating = rating_columns,
  Correlation_with_Demand = rating_demand_cors
), caption = "Correlation between Ratings and Demand Proxy")
```

**Reviews and Ratings Insights**: The distribution of reviews is highly skewed, with many properties having few reviews and a small number having hundreds of reviews. This is why we've used review counts as a key component in our demand proxy calculation. The ratings show a general trend toward high scores, with most properties being rated above 90 out of 100. The correlation analysis reveals interesting relationships between ratings, price, and demand, which can inform both pricing strategies for hosts and our predictive modeling approach.

## Correlation Analysis

Let's examine correlations between numeric variables to understand relationships.

```{r correlation_analysis}
# Select numeric columns for correlation analysis
numeric_columns <- train_data %>%
  dplyr::select_if(is.numeric) %>%
  # Remove ID columns and other non-meaningful numeric columns
  dplyr::select(-contains("id"), -contains("listing"), -contains("host_id"))

# Select a subset of most relevant columns for correlation analysis
key_columns <- c("price", "accommodates", "bathrooms", "bedrooms", "beds", 
                "reviews", "overall_rating", "demand_proxy", 
                "distance_to_center_km", "price_per_person", "host_experience_days")

# Filter to only include columns that exist
key_columns <- key_columns[key_columns %in% names(numeric_columns)]
key_numeric <- numeric_columns[, key_columns]

# Calculate correlation matrix
cor_matrix <- cor(key_numeric, use = "pairwise.complete.obs")

# Plot correlation matrix
corrplot(cor_matrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.7,
         title = "Correlation Matrix of Key Variables")

# Display top correlations with price
price_cors <- cor_matrix["price", ]
price_cors <- price_cors[order(abs(price_cors), decreasing = TRUE)]

cat("\nTop correlations with price:\n")
print(price_cors)

# Display top correlations with demand_proxy
demand_cors <- cor_matrix["demand_proxy", ]
demand_cors <- demand_cors[order(abs(demand_cors), decreasing = TRUE)]

cat("\nTop correlations with demand_proxy:\n")
print(demand_cors)
```

**Correlation Insights**: The correlation analysis reveals important relationships between variables. Price is strongly correlated with the number of accommodates, bedrooms, and bathrooms, as expected. Interestingly, our demand proxy shows strong correlation with reviews (which makes sense as it's part of the calculation) but less so with price, suggesting that demand isn't solely driven by pricing. The distance to the city center shows a negative correlation with price, confirming that central locations command higher prices. These relationships will be valuable for our predictive modeling.

## Key Insights Summary

Based on our exploratory data analysis, we've identified several key insights:

1. **Demand Distribution**: The demand proxy shows a right-skewed distribution, with most properties having low to moderate demand and fewer properties having very high demand.

2. **Price Patterns**:
   - Median price is around €49 per night
   - Prices follow a right-skewed distribution
   - Central neighborhoods have higher prices
   - Entire homes/apartments command higher prices than private rooms

3. **Geographic Patterns**:
   - Listings are concentrated in central districts
   - Both price and demand show clear spatial patterns
   - Distance to city center is negatively correlated with price

4. **Neighborhood Insights**:
   - Top neighborhoods by listing count: Neukölln, Friedrichshain, Prenzlauer Berg
   - Top neighborhoods by price aren't necessarily the same as by count
   - Certain neighborhoods show notably higher demand

5. **Property Characteristics**:
   - Apartments dominate the Berlin Airbnb market
   - Number of accommodates, bedrooms, and bathrooms are key price drivers
   - Room type significantly impacts both price and demand

6. **Rating Patterns**:
   - Most properties have high ratings (90+ out of 100)
   - Ratings show some correlation with demand but less with price

These insights will guide our feature selection and modeling approach in the next phase of the project.