---
title: "Berlin Airbnb Rental Demand Prediction - Exploratory Data Analysis"
author: "Data Science Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                     warning = FALSE, 
                     message = FALSE,
                     fig.width = 10, 
                     fig.height = 6)

# Load necessary libraries
library(tidyverse)    # For data manipulation and visualization
library(lubridate)    # For date manipulation
library(sf)           # For spatial data analysis
library(leaflet)      # For interactive maps
library(skimr)        # For data summary
library(corrplot)     # For correlation plots
library(GGally)       # For ggpairs plots
library(knitr)        # For nice tables
library(gridExtra)    # For grid arrangements
```

# Introduction

This document contains the exploratory data analysis (EDA) for the Airbnb Berlin rental demand prediction project. The goal is to gain insights from the data and identify patterns that will inform our predictive modeling approach.

## Data Loading

First, we need to load the dataset from the raw data directory.

```{r load_data}
# Set the file path
data_path <- "../data/raw/"

# List files in the directory - uncomment when data is available
# list.files(data_path)

# For now we'll create a placeholder message since data hasn't been downloaded yet
if (!file.exists(paste0(data_path, "airbnb_berlin.csv"))) {
  message("Data file not found! Please download the 'Airbnb Berlin Price Prediction' dataset from Kaggle and place it in the data/raw directory.")
}

# Placeholder for data loading code
# airbnb_data <- read.csv(paste0(data_path, "airbnb_berlin.csv"))

# Create a small sample data for demonstration purposes until real data is loaded
set.seed(123)
airbnb_data <- data.frame(
  Listing_ID = 1:10,
  Listing_Name = paste("Sample Listing", 1:10),
  Neighborhood = sample(c("Mitte", "Kreuzberg", "Neukölln", "Prenzlauer Berg"), 10, replace = TRUE),
  Property_Type = sample(c("Apartment", "House", "Room"), 10, replace = TRUE),
  Room_Type = sample(c("Entire home/apt", "Private room", "Shared room"), 10, replace = TRUE),
  Accommodates = sample(1:6, 10, replace = TRUE),
  Bathrooms = sample(c(1, 1.5, 2), 10, replace = TRUE),
  Bedrooms = sample(1:3, 10, replace = TRUE),
  Beds = sample(1:4, 10, replace = TRUE),
  Price = round(runif(10, 40, 200), 2),
  Reviews = sample(0:50, 10, replace = TRUE),
  Overall_Rating = round(runif(10, 3, 5), 1)
)

# Display sample data
kable(head(airbnb_data))
```

## Data Overview

Let's examine the structure and summary statistics of our data.

```{r data_summary}
# This will be replaced with real data exploration when dataset is available

# Structure of the data
str(airbnb_data)

# Summary statistics
summary(airbnb_data)

# Using skimr for a more comprehensive summary (will be used with full dataset)
# skim(airbnb_data)
```

# Planned Exploratory Analysis

Once the full dataset is available, we will conduct the following analyses:

## 1. Price Distribution Analysis

```{r price_distribution, eval=FALSE}
# Code for analyzing price distribution
ggplot(airbnb_data, aes(x = Price)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Airbnb Prices in Berlin",
       x = "Price (Euro)",
       y = "Count") +
  theme_minimal()

# Log transformation for skewed price data
ggplot(airbnb_data, aes(x = log(Price))) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Log-transformed Airbnb Prices in Berlin",
       x = "Log(Price)",
       y = "Count") +
  theme_minimal()

# Price statistics by neighborhood
ggplot(airbnb_data, aes(x = reorder(Neighborhood, Price, FUN = median), y = Price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Price Distribution by Neighborhood",
       x = "Neighborhood",
       y = "Price (Euro)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2. Geographic Analysis

```{r geographic_analysis, eval=FALSE}
# Code for geographic visualization
# This will involve creating maps with price/demand hotspots
# We'll use longitude and latitude data with leaflet or ggmap

# Example (placeholder):
# airbnb_sf <- st_as_sf(airbnb_data, coords = c("Longitude", "Latitude"), crs = 4326)
# 
# ggplot() +
#   geom_sf(data = berlin_districts) +
#   geom_sf(data = airbnb_sf, aes(color = Price), alpha = 0.6) +
#   scale_color_viridis_c() +
#   labs(title = "Airbnb Prices Across Berlin",
#        color = "Price (Euro)") +
#   theme_minimal()
```

## 3. Property and Host Features Analysis

```{r property_features, eval=FALSE}
# Code to analyze how property features relate to price and potentially demand
# Relationships between price and:
# - Property type
# - Room type
# - Number of accommodates
# - Bedrooms/bathrooms
# - Host experience
# - Ratings

# Example (placeholder):
ggplot(airbnb_data, aes(x = Accommodates, y = Price)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(title = "Relationship Between Capacity and Price",
       x = "Accommodates (number of people)",
       y = "Price (Euro)") +
  theme_minimal()

ggplot(airbnb_data, aes(x = Room_Type, y = Price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Price by Room Type",
       x = "Room Type",
       y = "Price (Euro)") +
  theme_minimal()
```

## 4. Correlation Analysis

```{r correlation_analysis, eval=FALSE}
# Code for correlation matrix of numeric variables
# Example (placeholder):
# numeric_data <- airbnb_data %>% select_if(is.numeric)
# correlation_matrix <- cor(numeric_data, use = "complete.obs")
# corrplot(correlation_matrix, method = "circle", type = "upper", 
#          tl.col = "black", tl.srt = 45)
```

## 5. Temporal Analysis

```{r temporal_analysis, eval=FALSE}
# Code to analyze time patterns if date data is available
# - Seasonality
# - Day of week effects
# - Holiday effects
```

# Creating a Demand Proxy

Since our dataset may not have a direct demand measure, we need to create one. Several approaches can be considered:

```{r demand_proxy, eval=FALSE}
# We'll develop and test different demand proxies based on:
# 1. Review frequency (reviews per month)
# 2. Price relative to neighborhood average
# 3. Booking rate (if available)
# 4. Composite index of multiple factors

# Placeholder example:
# airbnb_data <- airbnb_data %>%
#   mutate(
#     first_review_date = as.Date(First_Review),
#     last_review_date = as.Date(Last_Review),
#     days_on_platform = as.numeric(difftime(last_review_date, first_review_date, units = "days")),
#     reviews_per_month = ifelse(days_on_platform > 0, Reviews / (days_on_platform/30), 0),
#     demand_proxy = reviews_per_month
#   )
```

# Conclusion and Next Steps

This document has outlined the exploratory data analysis plan for the Airbnb Berlin rental demand prediction project. Once the actual dataset is acquired, we will:

1. Complete the EDA following this plan
2. Create appropriate demand proxies
3. Identify key features influencing demand
4. Prepare the data for predictive modeling

The next steps will involve feature engineering and model development, which will be documented in subsequent R Markdown files. 