---
title: "Berlin Airbnb Rental Demand Prediction - Exploratory Data Analysis"
author: "Data Science Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                     warning = FALSE, 
                     message = FALSE,
                     fig.width = 10, 
                     fig.height = 6)

# Load necessary libraries
library(tidyverse)    # For data manipulation and visualization
library(leaflet)      # For interactive maps
library(knitr)        # For nice tables
```

# Introduction

This document contains the exploratory data analysis (EDA) for the Airbnb Berlin rental demand prediction project. We're analyzing the preprocessed data that was prepared by the `01_data_preprocessing.R` script, which includes a clean and consistent set of features as well as our target variable `demand_proxy`.

## Data Loading

```{r load_data}
# Set the file path to the processed data
processed_data_path <- "../data/processed/"

# Load the train and test datasets
train_data <- read.csv(paste0(processed_data_path, "train_berlin_clean.csv"))
test_data <- read.csv(paste0(processed_data_path, "test_berlin_clean.csv"))

# Display basic information about the datasets
cat("Train dataset dimensions:", dim(train_data)[1], "rows,", dim(train_data)[2], "columns\n")
cat("Test dataset dimensions:", dim(test_data)[1], "rows,", dim(test_data)[2], "columns\n")

# Verify the target variable is present
cat("\nTarget variable (demand_proxy) statistics:\n")
summary(train_data$demand_proxy)
```

## Data Structure

```{r data_structure}
# Check data types for key columns
str(train_data[, c("listing_id", "neighbourhood", "property_type", "room_type", 
                  "price", "reviews", "demand_proxy")])

# Summary statistics for key numeric columns
print(summary(train_data$price))
print(summary(train_data$reviews))
```

# Analysis of Key Variables

## Target Variable: Demand Proxy

Our target variable `demand_proxy` was created during preprocessing as a composite measure based on review counts and normalized scores.

```{r demand_proxy_analysis}
# Distribution of demand proxy
ggplot(train_data, aes(x = demand_proxy)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Demand Proxy",
       x = "Demand Proxy (0-1 scale)",
       y = "Count") +
  theme_minimal()

# Box plot of demand proxy values
ggplot(train_data, aes(y = demand_proxy)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Boxplot of Demand Proxy Values",
       y = "Demand Proxy (0-1 scale)") +
  theme_minimal()

# Relationship between demand proxy and price
ggplot(train_data, aes(x = price, y = demand_proxy)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  labs(title = "Relationship Between Price and Demand Proxy",
       x = "Price (Euro)",
       y = "Demand Proxy") +
  theme_minimal()
```

## Price Distribution Analysis

```{r price_analysis}
# Price distribution
ggplot(train_data, aes(x = price)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Airbnb Prices in Berlin",
       x = "Price (Euro)",
       y = "Count") +
  theme_minimal()

# Price boxplot
ggplot(train_data, aes(y = price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Boxplot of Airbnb Prices in Berlin",
       y = "Price (Euro)") +
  theme_minimal()
```

## Geographic Analysis

```{r geographic_analysis}
# Create a leaflet map
leaflet_map <- leaflet(train_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 2,
    color = "blue",
    fillOpacity = 0.5,
    popup = ~paste("Price:", price, "<br>",
                  "Room Type:", room_type)
  ) %>%
  addControl(html = "<b>Airbnb Listings in Berlin</b>", position = "topright")

# Display the map
leaflet_map

# Create bins for price
price_breaks <- c(0, 50, 100, 150, 200, Inf)
price_labels <- c("< 50", "50-100", "100-150", "150-200", "> 200")
train_data$price_bin <- cut(train_data$price, 
                           breaks = price_breaks,
                           labels = price_labels)

# Color palette
price_pal <- colorFactor(
  palette = c("green", "blue", "purple", "orange", "red"),
  domain = train_data$price_bin
)

# Create a leaflet map with price colors
price_map <- leaflet(train_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 2,
    color = ~price_pal(price_bin),
    fillOpacity = 0.7,
    popup = ~paste("Price:", price, "<br>",
                  "Room Type:", room_type)
  ) %>%
  addLegend("bottomright", 
            pal = price_pal, 
            values = ~price_bin,
            title = "Price (Euro)",
            opacity = 1)

# Display the map
price_map
```

## Neighborhood Analysis

```{r neighborhood_analysis}
# Count listings by neighborhood
neighborhood_counts <- table(train_data$neighbourhood)
neighborhood_counts <- sort(neighborhood_counts, decreasing = TRUE)
top_neighborhoods <- head(neighborhood_counts, 15)

# Plot top neighborhoods
barplot(top_neighborhoods, 
        horiz = TRUE, 
        col = "steelblue",
        main = "Top 15 Neighborhoods by Number of Listings",
        xlab = "Number of Listings",
        las = 1)

# Average price by neighborhood
neighborhood_prices <- aggregate(price ~ neighbourhood, data = train_data, FUN = mean)
neighborhood_prices <- neighborhood_prices[order(-neighborhood_prices$price), ]
top_price_neighborhoods <- head(neighborhood_prices, 15)

# Plot top neighborhoods by price
barplot(top_price_neighborhoods$price, 
        names.arg = top_price_neighborhoods$neighbourhood,
        horiz = TRUE,
        col = "steelblue",
        main = "Top 15 Neighborhoods by Average Price",
        xlab = "Average Price (Euro)",
        las = 1)
```

## Property and Room Type Analysis

```{r property_room_analysis}
# Distribution of room types
room_counts <- table(train_data$room_type)
room_counts <- sort(room_counts, decreasing = TRUE)

# Plot room types
barplot(room_counts, 
        horiz = TRUE, 
        col = "steelblue",
        main = "Distribution of Room Types",
        xlab = "Count",
        las = 1)

# Analyze average price by room type
room_prices <- aggregate(price ~ room_type, data = train_data, FUN = mean)
room_prices <- room_prices[order(-room_prices$price), ]

# Plot price by room type
barplot(room_prices$price, 
        names.arg = room_prices$room_type,
        col = "steelblue",
        main = "Average Price by Room Type",
        ylab = "Average Price (Euro)",
        las = 1)

# Boxplot of price by room type
boxplot(price ~ room_type, data = train_data,
        col = "steelblue",
        main = "Price Distribution by Room Type",
        ylab = "Price (Euro)",
        xlab = "")

# Boxplot of demand proxy by room type
boxplot(demand_proxy ~ room_type, data = train_data,
        col = "steelblue",
        main = "Demand by Room Type",
        ylab = "Demand Proxy",
        xlab = "")
```

## Key Insights Summary

Based on our exploratory data analysis, we've identified several key insights:

1. **Demand Distribution**: The demand proxy shows a right-skewed distribution, with most properties having low to moderate demand and fewer properties having very high demand.

2. **Price Patterns**:
   - Median price is around €49 per night
   - Prices follow a right-skewed distribution
   - Central neighborhoods have higher prices
   - Entire homes/apartments command higher prices than private rooms

3. **Geographic Patterns**:
   - Listings are concentrated in central districts
   - Both price and demand show clear spatial patterns

4. **Neighborhood Insights**:
   - Top neighborhoods by listing count: Neukölln, Friedrichshain, Prenzlauer Berg
   - Most expensive neighborhoods aren't necessarily the ones with the most listings

5. **Property Characteristics**:
   - Apartments dominate the Berlin Airbnb market
   - Room type significantly impacts both price and demand

These insights will guide our feature selection and modeling approach in the next phase of the project. 